Sensor Clusterization in D-optimal Design in Infinite Dimensional Bayesian Linear Inverse Problems

### 2020-10-26 new_paper.tex
Cite karina and ahmed, changed propostions, theorems, lemmas to upper/lowercase, discarded the requirement for zero mean

\begin{abstract}
  We investigate the problem of sensor clusterization in optimal
  experimental design for infinite-dimensional Bayesian linear inverse
  problems. We suggest an analytically tractable model for such
  designs and reason how it may lead to sensor clusterization in the
  case of iid measurement noise. We also show that in the case of
  spatially correlated measurement error clusterization does not
  occur. As a part of the analysis we prove a matrix determinant lemma
  analog in infinite dimensions and a lemma for calculating
  $\frac{\der \log \det (I + X(t))}{\der t}$ for certain
  operator-valued functions. We also show how to decompose $M = AA^t$
  with $A$ constrained to have unit norm columns.
\end{abstract}

@@

@@

A beautiful mathematical structure arises in D-optimal designs when no
model error is present. The Lagrange multipliers problems is in fact a
nonlinear eigenvalue problem for the observations and prior
covariance. The operator for which eigenvectors and eigenvalues are
sought is a sum of two operators. The first is the prior
covariance. The second is an outer product of the observations (see
\eqref{eq:eigenproblem} for details and exact statement). This
structure helps us shed light into D-optimal designs. We characterize
these designs using theorem \ref{thm:char}. One insight we prove in
said theorem, is that a D-optimal design reduces uncertainties where
they are highest first. Other interestig phenomena arise, but they
require setting notation, and are discussed later.

In the process we generalize several lemmas from linear algebra to
infinite-dimensional settings. We prove a matrix determinant lemma in
\ref{lemma:MDL}. We generalize a lemma due to Lax \cite{lax97} for
calculating $\frac{\der}{\der t} \log \det (I + X(t))$, for an
operator valued function $X(t)$ in \ref{lemma:lax}. We also prove
lemmas in linear algebra. One gives a decomposition $M = AA^t$ where
$A$ has unit norm columns \ref{lemma:free}. Another shows simultaneous
diagonizability of the prior and outer product in the nonlinear
eigenvalue problem referenced above \ref{lemma:sim diag}. We also
provide other tools for understanding D-optimal designs in
infinite-dimensional Bayesian inverse problems, including lemmas for
calculating the increase in the design criterion per observation ---
Lemmas \ref{lemma:design increase} and Corollary \ref{cor:zero mod
  err}.

@@

\subsection{Bayesian Linear Inverse Problems}\label{subsec:abstract OED}
Let $\hilp$ and $\hilo$ be separable Hilbert spaces (the subscripts p
and o are for ``parameter'' and ``observation'', respectfully). We
denote $\hilo^*$ the Hilbert space of all linear functionals on
$\hilo$. Consider $\fwd: \hilp \to \hilo$, a linear operator (the
``forward operator''). We are interested in forward operators that are
strongly smoothing (have fast decaying modes --- the heat operator
from section \ref{subsec:example} is a prime example). Our goal is to
infer $\param \in \hilp$ --- some parameter of the dynamics --- given
noisy observations of $\fwd \param$. We take a Gaussian prior $\pr$
for $\param$. Hence, $\param \sim \pr = \normal(\param_0 ,\prcov)$
with some appropriate covariance operator $\prcov$ \cite{Stuart10}. It
is important to note that $\fwd \prcov \fwd^*$ is the prior covariance
in $\hilo$ \cite{Stuart10}. As such, it is assumed invertible --- an
assumption we rely on below. Throughout this paper, we denote $m$ the
number of measurements we are allowed to take. Observations
(measurements) are taken via the linear observation operator $\obs \in
( \hilo^* )^m$. In an analogy with linear algebra (where row vectors
are thought of as linear functionals), we think of $\obs$ as having
``rows'' $\meas_j, j=1,\dots,m$:
\begin{equation}\label{eq:O}
  \obs = (\meas_1,\dots, \meas_m)^t, \meas_j \in \hilo^*, j = 1,\dots,m.
\end{equation}
This way, for $u \in \hilo$ we have $\obs u = (\meas_1(u), \dots,
\meas_m(u) )^t \in \R^m$.
%% but we may drop the parentheses and write $\meas_j u = \meas_j(u)$.
%% A few observations regarding $\obs$ are in order.
%% First, it is good to keep in mind that $( \hilo^* )^m$ is a Hilbert
%% space with norm $\| \obs \| = \sum_{j=1}^m \|\meas_j\|$.
For $u \in \hilo$ and $v\in \R^m$:
\begin{align*}
  \big (\obs^*v \big ) (u) &= \langle v, \obs u \rangle_{\R^m} = \sum_{j=1}^m  v_j \meas_j(u)
  = v^t \left ( \obs u \right ) = (v^t \obs) (u),
\end{align*}
and thus:
\begin{align}\label{eq:obs*}
  \obs^*v &= \sum_{j=1}^m v_j \meas_j = v^t \obs.
\end{align}

@@

The meaning of D-optimal design in infinite-dimensional Hilbert spaces
was investigated in \cite{AlexanderianGloorGhattas14}. There, the
authors make assumptions that amount to $\Sigma=I$ (implied by
$\modcov = 0,\sigma^2=1$), but we choose not to take these
simplification here. This is because $\Sigma$ can determine ``how
much'' clusterization we see. As stated
\cite[pp. 681]{AlexanderianGloorGhattas14}, the results hold for more
general covariance matrices. The conclusion is that in
infinite-dimensions, a D-optimal design is well-defined as maximizing
the expected KL divergence between posterior and prior. The main
result is summarized, using our notation, in the following theorem:
\begin{theorem}[Slightly modified Theorem 1 from \cite{AlexanderianGloorGhattas14}]
  Let $\pr = \normal(\param_0,\prcov)$ be a Gaussian prior on $\hilp$
  and let $\post = \normal(\postmean,\postcov)$ the posterior measure
  on $\hilo$ for the Bayesian linear inverse problem $\data = \obs
  \fwd\param + \obs \eps' + \eps$ discussed above. Then
  \begin{align}\label{eq:objective}
    \begin{split}
      \tar( \obs) :&= \mathbb{E}_{\data}\left [ D_{\text{KL}} (\post || \pr ) \right ] \\
      % 
      % 
      % 
      &= \frac12 \log \det 
      ( I + \prcov^{1/2}  \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}).
    \end{split}
  \end{align}
\end{theorem}
\begin{definition}\label{def:d optimality}
  A design $\obs^{\star}$ is said to be D-optimal if $\obs^{\star} = \argmax_{\obs} \tar(\obs)$.
\end{definition}

@@

Below, the expression for the posterior covariance \eqref{eq:postcov}
justifies the second equality. Some calculations delegated to lemma
\ref{lemma:twice woodbury} in the appendix justify the third (trivial
if $\fwd$ is invertible):
\begin{align}\label{eq:mod conditions}
  \begin{split}
  \obs^* \Xi &= \sigma^{-2}\fwd \postcov \fwd^* \obs^*  \\
  %
  %
  %
  &= \sigma^{-2} \fwd( \prcov^{-1} + \sigma^{-2}  \fwd^* \obs^* \obs \fwd )^{-1} \fwd^* \obs^* \\
  %
  %
  %
  &= \sigma^{-2} \left ( (\fwd\prcov\fwd^*)^{-1} + \sigma^{-2}  \obs^* \obs \right )^{-1} \obs^*
  \end{split}
\end{align}  
The term $(\fwd \prcov \fwd^*)^{-1}$ is the prior precision in $\hilo$
(and is invertible, as mentioned in section
\ref{section:prelim}). $(\fwd\prcov\fwd^*)^{-1} + \sigma^{-2} \obs^*
\obs$ is the posterior precision in $\hilo$.

@@

\begin{corollary}[Eigenvectors of a D-optimal design]\label{cor:same ev}
  $\obs^*\obs$ has the same eigenvectors as $\fwd \prcov \fwd^*$.
  %namely $\{\ev_i\}_{i=1}^{\infty}$.
\end{corollary}
\begin{proof}
In \eqref{eq:mod conditions}, take $\func_j^{*} = \meas_j$ and $C :=
(\fwd \prcov \fwd^*)^{-1}$ and use lemma \ref{lemma:sim diag}.
\end{proof}
At most $m$ of the eigenvectors of $\obs^*\obs$ have positive
eigenvalue, because $k := \rank \obs^*\obs \leq m$. Denote the
corresponding non-zero eigenvalues of $\obs^*\obs$ by
$\{\eta_i\}_{i=1}^{k}$ and let $\eta_i = 0$ for $i \geq k+1$. No
assumption is made (yet) regarding the magnitudes of the corresponding
$\{\lambda_i\}_{i=1}^{\infty}$. They are to be thought of as fixed but
unknown. Using Sylvester's determinant theorem (the standard proof
works for Hilbert spaces as well) we get:

@@

One can understand D-optimal designs as follows: Imagine each
eigenvector corresponds to a graduated lab cylinder (figure
\ref{fig:cylinder}). Each cylinder $i$ is filled, a-priori, with green
liquid of $\sigma^2\lambda_i^{-1}$ volume units. This liquid
represnets the prior precision, weighted by observation noise. The
volume of the green liquid is increasing in $i$. A D-optimal design
with $m$ measurements ignores cylinders indexed $m+1$ and above. It
has blue liquid of $m$ volume units at its disposal. It distributes
said liquid by repeatedly adding a drop to whatever cylinder currently
has the lowest level of liquid in it (lowest precision). The result of
such a procedure is illustrated in figure \ref{fig:optimal vs
  not}. The analysis conducted in this section is summarized in
theorem \ref{thm:char} below.

@@

\section{Sensor Clusterization --- Vanishing Model Error}\label{section:clusterization}
\subsection{Overview}
In this section we use theorem \ref{thm:char} to explain how sensor
clusterization may arise in various scenarios. We first consider
sequential designs, and then simultaneous designs.

@@

\subsection{Simultaneous Design}\label{subsec:clusterization simultaneous}
In a simultaneous design scenario, the result is not as clear cut as
in the sequential design case. We still rely on previous results ---
namely corollary \ref{cor:same ev} and theorem \ref{thm:char}. An
illustration of two D-optimal designs scenarios for the same problem
are presented in figure \ref{fig:clusterization} --- the first
exhibits clusterization and second does not. As was discussed earlier
and illustrated in figure \ref{fig:optimal vs not}, an optimal design
makes the precision ``flat'' where it is lowest. Clusterization may
arise when two or more eigenvectors have equal precision. This
happens, for example for eigenvectors $1$ and $2$ when considering
measurements $\meas_i,i=1,2,3$. Taking three measurements, it is
possible that $\meas_1$ is such that it equalizes the precision of
eigenvectors $1$ and $2$. In this case, taking mesurements $\meas_2 =
\meas_3 = \frac{1}{2}(\ev_1 + \ev_2)$ leads to an optimal design that
exhibits clusterization. Similarly, taking $\meas_5 = \meas_6 =
\frac{1}{3}(\ev_1 + \ev_2 + \ev_3)$ also gives rise to an optimal
design that exhibits clusterization. It is important to note that
there are (infinitely) many optimal designs that actually do not
exhibit clusterization --- e.g. the second design presented in figure
\ref{fig:clusterization}.

>>>>

\begin{abstract}
  We investigate the problem of sensor clusterization in optimal
  experimental design for infinite-dimensional Bayesian linear inverse
  problems. We suggest an analytically tractable model for such
  designs and reason how it may lead to sensor clusterization in the
  case of iid measurement noise. We also show that in the case of
  spatially correlated measurement error clusterization does not
  occur. As a part of the analysis we prove a Matrix Determinant Lemma
  analog in infinite dimensions and a lemma for calculating
  $\frac{\der \log \det (I + X(t))}{\der t}$ for certain
  operator-valued functions. We also show how to decompose $M = AA^t$
  with $A$ constrained to have unit norm columns.
\end{abstract}

@@

In the inverse problems community, work is mostly computational and
less theoretic. Model errors were considered in \cite{Attia20,
  Koval20}. The former study is focused on inferring of a Quantity of
Interest (QoI). The focus of the latter is reducing forward solves,
using randomized linear algebra. Both studies present numerical
techniques for finding optimal designs when model error is
present. Both are restricted to linear inverse problems (although in
the latter the authors use their method on a nonlinear problem with a
Laplace approximation for the posterior). Both find an optimal design
by first solving a continuous problem for sensor weights. Said
solution is then sparsified to give a binary design. Both studies are
successful in the task of Bayesian inversion. However, neither of
these studies mentions any effect model errors can have on sensor
clusterization. The current study is mostly theoretical and aims to
fill the theoretical gap of understanding sensor clusterization.

@@

A beautiful mathematical structure arises in D-optimal designs when no
model error is present. The Lagrange multipliers problems is in fact a
nonlinear eigenvalue problem for the observations and prior
covariance. The operator for which eigenvectors and eigenvalues are
sought is a sum of two operators. The first is the prior
covariance. The second is an outer product of the observations (see
\eqref{eq:eigenproblem} for details and exact statement). This
structure helps us shed light into D-optimal designs. We characterize
these designs using Theorem \ref{thm:char}. One insight we prove in
said theorem, is that a D-optimal design reduces uncertainties where
they are highest first. Other interestig phenomena arise, but they
require setting notation, and are discussed later.

In the process we generalize several lemmas from linear algebra to
infinite-dimensional settings. We prove a Matrix Determinant Lemma in
\ref{lemma:MDL}. We generalize a lemma due to Lax \cite{lax97} for
calculating $\frac{\der}{\der t} \log \det (I + X(t))$, for an
operator valued function $X(t)$ in \ref{lemma:lax}. We also prove
lemmas in linear algebra. One gives a decomposition $M = AA^t$ where
$A$ has unit norm columns \ref{lemma:free}. Another shows simultaneous
diagonizability of the prior and outer product in the nonlinear
eigenvalue problem referenced above \ref{lemma:sim diag}. We also
provide other tools for understanding D-optimal designs in
infinite-dimensional Bayesian inverse problems, including lemmas for
calculating the increase in the design criterion per observation ---
Lemma \ref{lemma:design increase} and Corollary \ref{cor:zero mod
  err}.

@@

\subsection{Bayesian Linear Inverse Problems}\label{subsec:abstract OED}
Let $\hilp$ and $\hilo$ be separable Hilbert spaces (the subscripts p
and o are for ``parameter'' and ``observation'', respectfully). We
denote $\hilo^*$ the Hilbert space of all linear functionals on
$\hilo$. Consider $\fwd: \hilp \to \hilo$, a linear operator (the
``forward operator''). We are interested in forward operators that are
strongly smoothing (have fast decaying modes --- the heat operator
from section \ref{subsec:example} is a prime example). Our goal is to
infer $\param \in \hilp$ --- some parameter of the dynamics --- given
noisy observations of $\fwd \param$. We take a Gaussian prior $\pr$
for $\param$. Hence, $\param \sim \pr = \normal(\prmean ,\prcov)$
with some appropriate covariance operator $\prcov$ \cite{Stuart10}. It
is important to note that $\fwd \prcov \fwd^*$ is the prior covariance
in $\hilo$ \cite{Stuart10}. As such, it is assumed invertible --- an
assumption we rely on below. Throughout this paper, we denote $m$ the
number of measurements we are allowed to take. Observations
(measurements) are taken via the linear observation operator $\obs \in
( \hilo^* )^m$. In an analogy with linear algebra (where row vectors
are thought of as linear functionals), we think of $\obs$ as having
``rows'' $\meas_j, j=1,\dots,m$:
\begin{equation}\label{eq:O}
  \obs = (\meas_1,\dots, \meas_m)^t, \meas_j \in \hilo^*, j = 1,\dots,m.
\end{equation}
This way, for $u \in \hilo$ we have $\obs u = (\meas_1(u), \dots,
\meas_m(u) )^t \in \R^m$.
%% but we may drop the parentheses and write $\meas_j u = \meas_j(u)$.
%% A few observations regarding $\obs$ are in order.
%% First, it is good to keep in mind that $( \hilo^* )^m$ is a Hilbert
%% space with norm $\| \obs \| = \sum_{j=1}^m \|\meas_j\|$.
For $u \in \hilo$ and $v\in \R^m$:
\begin{align*}
  \big (\obs^*v \big ) (u) &= \langle v, \obs u \rangle_{\R^m} = \sum_{j=1}^m  v_j \meas_j(u)
  = v^t \left ( \obs u \right ) = (v^t \obs) (u),
\end{align*}
and thus:
\begin{align}\label{eq:obs*}
  \obs^*v &= \sum_{j=1}^m v_j \meas_j = v^t \obs.
\end{align}

@@

The meaning of D-optimal design in infinite-dimensional Hilbert spaces
was investigated in \cite{AlexanderianGloorGhattas14}. There, the
authors make assumptions that amount to $\Sigma=I$ (implied by
$\modcov = 0,\sigma^2=1$), but we choose not to take these
simplification here. This is because $\Sigma$ can determine ``how
much'' clusterization we see. As stated
\cite[pp. 681]{AlexanderianGloorGhattas14}, the results hold for more
general covariance matrices. The conclusion is that in
infinite-dimensions, a D-optimal design is well-defined as maximizing
the expected KL divergence between posterior and prior. The main
result is summarized, using our notation, in the following theorem:
\begin{theorem}[Slightly modified Theorem 1 from \cite{AlexanderianGloorGhattas14}]
  Let $\pr = \normal(\prmean,\prcov)$ be a Gaussian prior on $\hilp$
  and let $\post = \normal(\postmean,\postcov)$ the posterior measure
  on $\hilo$ for the Bayesian linear inverse problem $\data = \obs
  \fwd\param + \obs \eps' + \eps$ discussed above. Then
  \begin{align}\label{eq:objective}
    \begin{split}
      \tar( \obs) :&= \mathbb{E}_{\data}\left [ D_{\text{KL}} (\post || \pr ) \right ] \\
      % 
      % 
      % 
      &= \frac12 \log \det 
      ( I + \prcov^{1/2}  \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}).
    \end{split}
  \end{align}
\end{theorem}
\begin{definition}\label{def:d optimality}
  A design $\obs^{\star}$ is said to be D-optimal if $\obs^{\star} = \argmax_{\obs} \tar(\obs)$.
\end{definition}

@@

Below, the expression for the posterior covariance \eqref{eq:postcov}
justifies the second equality. Some calculations delegated to Lemma
\ref{lemma:twice woodbury} in the appendix justify the third (trivial
if $\fwd$ is invertible):
\begin{align}\label{eq:mod conditions}
  \begin{split}
  \obs^* \Xi &= \sigma^{-2}\fwd \postcov \fwd^* \obs^*  \\
  %
  %
  %
  &= \sigma^{-2} \fwd( \prcov^{-1} + \sigma^{-2}  \fwd^* \obs^* \obs \fwd )^{-1} \fwd^* \obs^* \\
  %
  %
  %
  &= \sigma^{-2} \left ( (\fwd\prcov\fwd^*)^{-1} + \sigma^{-2}  \obs^* \obs \right )^{-1} \obs^*
  \end{split}
\end{align}  
The term $(\fwd \prcov \fwd^*)^{-1}$ is the prior precision in $\hilo$
(and is invertible, as mentioned in section
\ref{section:prelim}). $(\fwd\prcov\fwd^*)^{-1} + \sigma^{-2} \obs^*
\obs$ is the posterior precision in $\hilo$.

@@

\begin{corollary}[Eigenvectors of a D-optimal design]\label{cor:same ev}
  $\obs^*\obs$ has the same eigenvectors as $\fwd \prcov \fwd^*$.
  %namely $\{\ev_i\}_{i=1}^{\infty}$.
\end{corollary}
\begin{proof}
In \eqref{eq:mod conditions}, take $\func_j^{*} = \meas_j$ and $C :=
(\fwd \prcov \fwd^*)^{-1}$ and use Lemma \ref{lemma:sim diag}.
\end{proof}
At most $m$ of the eigenvectors of $\obs^*\obs$ have positive
eigenvalue, because $k := \rank \obs^*\obs \leq m$. Denote the
corresponding non-zero eigenvalues of $\obs^*\obs$ by
$\{\eta_i\}_{i=1}^{k}$ and let $\eta_i = 0$ for $i \geq k+1$. No
assumption is made (yet) regarding the magnitudes of the corresponding
$\{\lambda_i\}_{i=1}^{\infty}$. They are to be thought of as fixed but
unknown. Using Sylvester's Determinant Theorem (the standard proof
works for Hilbert spaces as well) we get:

@@

One can understand D-optimal designs as follows: Imagine each
eigenvector corresponds to a graduated lab cylinder (figure
\ref{fig:cylinder}). Each cylinder $i$ is filled, a-priori, with green
liquid of $\sigma^2\lambda_i^{-1}$ volume units. This liquid
represnets the prior precision, weighted by observation noise. The
volume of the green liquid is increasing in $i$. A D-optimal design
with $m$ measurements ignores cylinders indexed $m+1$ and above. It
has blue liquid of $m$ volume units at its disposal. It distributes
said liquid by repeatedly adding a drop to whatever cylinder currently
has the lowest level of liquid in it (lowest precision). The result of
such a procedure is illustrated in figure \ref{fig:optimal vs
  not}. The analysis conducted in this section is summarized in
Theorem \ref{thm:char} below.

@@

\section{Sensor Clusterization --- Vanishing Model Error}\label{section:clusterization}
\subsection{Overview}
In this section we use Theorem \ref{thm:char} to explain how sensor
clusterization may arise in various scenarios. We first consider
sequential designs, and then simultaneous designs.

@@

\subsection{Simultaneous Design}\label{subsec:clusterization simultaneous}
In a simultaneous design scenario, the result is not as clear cut as
in the sequential design case. We still rely on previous results ---
namely Corollary \ref{cor:same ev} and Theorem \ref{thm:char}. An
illustration of two D-optimal designs scenarios for the same problem
are presented in figure \ref{fig:clusterization} --- the first
exhibits clusterization and second does not. As was discussed earlier
and illustrated in figure \ref{fig:optimal vs not}, an optimal design
makes the precision ``flat'' where it is lowest. Clusterization may
arise when two or more eigenvectors have equal precision. This
happens, for example for eigenvectors $1$ and $2$ when considering
measurements $\meas_i,i=1,2,3$. Taking three measurements, it is
possible that $\meas_1$ is such that it equalizes the precision of
eigenvectors $1$ and $2$. In this case, taking mesurements $\meas_2 =
\meas_3 = \frac{1}{2}(\ev_1 + \ev_2)$ leads to an optimal design that
exhibits clusterization. Similarly, taking $\meas_5 = \meas_6 =
\frac{1}{3}(\ev_1 + \ev_2 + \ev_3)$ also gives rise to an optimal
design that exhibits clusterization. It is important to note that
there are (infinitely) many optimal designs that actually do not
exhibit clusterization --- e.g. the second design presented in figure
\ref{fig:clusterization}.
